---
title: "From a CES Benchmark to the AW Estimator"
author: "Child Skill Formation"
date: "2026-09-01"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

set.seed(20260901)
options(width = 100)
library(knitr)
```

# Purpose

This notebook is a **didactic implementation**, not a replication of
Agostinelli and Wiswall. It begins with a simple CES production function
where the latent inputs are observed. It then introduces measurement error
and finishes with a small translog-IV exercise inspired by AW.

The basic sequence is:

1. Estimate a CES benchmark.
2. Add noisy measures of latent skills and investments.
3. Show the measurement-error problem.
4. Use alternative measures as internal instruments.
5. Estimate a translog skill-formation equation.

# 1. CES benchmark with observed inputs

Consider a one-skill, one-investment CES technology:

$$
\theta_{t+1}
=
A
\left[
\gamma\theta_t^\rho
+(1-\gamma)I_t^\rho
\right]^{1/\rho}
\exp(\eta_{t+1}).
$$

Here, \(\rho\) controls the curvature of the CES aggregator. The limiting
case \(\rho\rightarrow 0\) is Cobb--Douglas, while \(\rho=1\) gives perfect
substitutes.

## Simulate latent data

```{r simulate-ces}
n <- 4000

theta_t <- exp(rnorm(n, mean = 0, sd = 0.70))

log_I_t <- 0.20 +
  0.40 * log(theta_t) +
  rnorm(n, mean = 0, sd = 0.60)

I_t <- exp(log_I_t)

A_true <- 1.10
gamma_true <- 0.65
rho_true <- -0.50

eta_t1 <- rnorm(n, mean = 0, sd = 0.08)

theta_t1 <- A_true *
  (
    gamma_true * theta_t^rho_true +
      (1 - gamma_true) * I_t^rho_true
  )^(1 / rho_true) *
  exp(eta_t1)

ces_data <- data.frame(
  theta_t = theta_t,
  I_t = I_t,
  theta_t1 = theta_t1
)
```

## Estimate the CES parameters

Because the latent inputs are observed in this first exercise, we can estimate
the production function directly by nonlinear least squares.

```{r estimate-ces}
ces_fit <- nls(
  theta_t1 ~ A *
    (
      gamma * theta_t^rho +
        (1 - gamma) * I_t^rho
    )^(1 / rho),
  data = ces_data,
  start = list(A = 1.0, gamma = 0.60, rho = -0.40),
  algorithm = "port",
  lower = c(A = 0.01, gamma = 0.001, rho = -5),
  upper = c(A = 10.0, gamma = 0.999, rho = 0.99)
)

ces_estimates <- data.frame(
  Parameter = c("A", "gamma", "rho"),
  True = c(A_true, gamma_true, rho_true),
  Estimate = coef(ces_fit)
)

kable(ces_estimates, digits = 3)
```

```{r ces-fit-plot}
ces_data$fitted <- predict(ces_fit)

plot(
  log(ces_data$theta_t1),
  log(ces_data$fitted),
  pch = 16,
  col = rgb(35, 78, 112, 90, maxColorValue = 255),
  xlab = "Log true next-period skill",
  ylab = "Log fitted next-period skill",
  main = "CES benchmark: fitted versus simulated outcome"
)
abline(0, 1, lwd = 2, col = "firebrick")
```

At this point, the exercise is only about the production function. There is
no latent-factor problem and no IV problem yet.

# 2. Add measurement error

Now suppose that skill and investment are latent. We observe several noisy
measures:

$$
Z_{\theta,m}
=
\mu_{\theta,m}
+
\lambda_{\theta,m}\ln\theta_t
+
\varepsilon_{\theta,m},
$$

$$
Z_{I,m}
=
\mu_{I,m}
+
\lambda_{I,m}\ln I_t
+
\varepsilon_{I,m}.
$$

For the demonstration, the first measure of each factor is normalized to
have loading one. The measurement errors are generated independently across
measures.

## Simulate latent variables and noisy measures

```{r simulate-measures}
n <- 4000

log_theta_0 <- rnorm(n, mean = 0, sd = 0.70)

log_I_0 <- 0.25 +
  0.35 * log_theta_0 +
  rnorm(n, mean = 0, sd = 0.50)

# A translog skill-formation technology used for the AW part.
b0_true <- 0.10
b1_true <- 0.55
b2_true <- 0.25
b3_true <- -0.10

log_theta_1 <- b0_true +
  b1_true * log_theta_0 +
  b2_true * log_I_0 +
  b3_true * log_theta_0 * log_I_0 +
  rnorm(n, mean = 0, sd = 0.20)

make_measures <- function(
    x,
    loadings,
    error_sd,
    prefix
) {
  out <- sapply(
    seq_along(loadings),
    function(j) {
      loadings[j] * x +
        rnorm(length(x), mean = 0, sd = error_sd[j])
    }
  )
  out <- as.data.frame(out)
  names(out) <- paste0(prefix, seq_along(loadings))
  out
}

theta_0_measures <- make_measures(
  x = log_theta_0,
  loadings = c(1.00, 0.80, 1.20),
  error_sd = c(0.35, 0.45, 0.30),
  prefix = "Z_theta0_"
)

I_0_measures <- make_measures(
  x = log_I_0,
  loadings = c(1.00, 0.90, 1.15),
  error_sd = c(0.40, 0.50, 0.35),
  prefix = "Z_I0_"
)

theta_1_measures <- make_measures(
  x = log_theta_1,
  loadings = c(1.00, 0.85, 1.10),
  error_sd = c(0.35, 0.45, 0.30),
  prefix = "Z_theta1_"
)

measure_data <- cbind(
  theta_0_measures,
  I_0_measures,
  theta_1_measures
)
```

## Recover loadings from covariance ratios

With three measures and an anchor loading normalized to one, the covariance
restrictions are

$$
\operatorname{Cov}(Z_1,Z_2)
=
\lambda_2\operatorname{Var}(X),
$$

$$
\operatorname{Cov}(Z_1,Z_3)
=
\lambda_3\operatorname{Var}(X),
$$

$$
\operatorname{Cov}(Z_2,Z_3)
=
\lambda_2\lambda_3\operatorname{Var}(X).
$$

For example,

$$
\lambda_2
=
\frac{\operatorname{Cov}(Z_2,Z_3)}
{\operatorname{Cov}(Z_1,Z_3)}.
$$

The following function implements this simple covariance-ratio calculation.

```{r recover-factor}
recover_factor <- function(z, anchor = 1) {
  z <- as.data.frame(z)
  C <- cov(z)
  m <- ncol(z)

  loadings <- rep(NA_real_, m)
  loadings[anchor] <- 1

  non_anchor <- setdiff(seq_len(m), anchor)

  for (j in non_anchor) {
    comparison_measure <- setdiff(non_anchor, j)[1]
    loadings[j] <-
      C[j, comparison_measure] /
      C[anchor, comparison_measure]
  }

  factor_variance <- mean(
    C[anchor, non_anchor] / loadings[non_anchor]
  )

  factor_mean <- colMeans(z)

  measurement_variance <- pmax(
    diag(C) - loadings^2 * factor_variance,
    0
  )

  residual_measures <- sweep(
    sweep(as.matrix(z), 2, factor_mean, "-"),
    2,
    loadings,
    "/"
  )

  residual_measures <- as.data.frame(residual_measures)
  names(residual_measures) <- names(z)

  list(
    loadings = loadings,
    factor_variance = factor_variance,
    factor_mean = factor_mean,
    measurement_variance = measurement_variance,
    residual_measures = residual_measures
  )
}

theta_0_factor <- recover_factor(theta_0_measures)
I_0_factor <- recover_factor(I_0_measures)

loading_table <- rbind(
  CognitiveSkill = theta_0_factor$loadings,
  Investment = I_0_factor$loadings
)

colnames(loading_table) <- c("Measure 1", "Measure 2", "Measure 3")
kable(loading_table, digits = 3)
```

The recovered residual measure has the form

$$
\widetilde Z_{\theta,m}
=
\ln\theta_t
+
\frac{\varepsilon_{\theta,m}}{\lambda_{\theta,m}}.
$$

It is therefore a better-scaled proxy, but it is still noisy.

# 3. Why naive OLS is problematic

Consider the investment equation

$$
\ln I_i
=
a_0+a_1\ln\theta_i+\eta_i.
$$

If we regress a noisy skill proxy on a noisy investment proxy, the
measurement error in the skill proxy enters the regression residual.
Consequently, the regressor is correlated with the regression error.

```{r naive-investment}
theta_proxy <- theta_0_factor$residual_measures$Z_theta0_1
theta_alternative <- theta_0_factor$residual_measures$Z_theta0_2

investment_proxy <- I_0_factor$residual_measures$Z_I0_1

naive_investment_fit <- lm(
  investment_proxy ~ theta_proxy
)

# A simple repeated-measure IV:
# use an alternative skill measure as an instrument.
investment_first_stage <- lm(
  theta_proxy ~ theta_alternative
)

theta_hat <- fitted(investment_first_stage)

iv_investment_fit <- lm(
  investment_proxy ~ theta_hat
)

investment_comparison <- data.frame(
  Method = c("True", "Naive OLS", "Repeated-measure IV"),
  Coefficient_on_skill = c(
    0.35,
    coef(naive_investment_fit)[["theta_proxy"]],
    coef(iv_investment_fit)[["theta_hat"]]
  )
)

kable(investment_comparison, digits = 3)
```

The internal IV works because the alternative measure is correlated with the
true skill but, under the measurement-error assumptions, is uncorrelated with
the measurement error in the main skill measure.

This is not an external policy IV. It is a repeated-measure IV for an
errors-in-variables problem.

# 4. A small AW-style translog exercise

The AW translog technology can be written as

$$
\ln\theta_{t+1}
=
\beta_{0,t}
+
\beta_{1,t}\ln\theta_t
+
\beta_{2,t}\ln I_t
+
\beta_{3,t}
\ln\theta_t\ln I_t
+
u_{t+1}.
$$

The coefficient \(\beta_{3,t}\) is the log cross-partial:

$$
\frac{\partial^2\ln\theta_{t+1}}
{\partial\ln\theta_t\,\partial\ln I_t}
=
\beta_{3,t}.
$$

In the simulated data, the true parameter values are:

```{r true-aw-parameters}
true_aw <- data.frame(
  Parameter = c(
    "Intercept",
    "Current skill",
    "Investment",
    "Skill-investment interaction"
  ),
  True = c(b0_true, b1_true, b2_true, b3_true)
)

kable(true_aw, digits = 3)
```

## Construct noisy regressors and instruments

For this simple demonstration:

- the main skill proxy is \(Z_{\theta,1}\);
- the main investment proxy is \(Z_{I,1}\);
- the alternative skill and investment measures are \(Z_{\theta,2}\)
  and \(Z_{I,2}\);
- the interaction of the alternative measures instruments the interaction
  of the main measures.

```{r construct-aw-data}
aw_data <- data.frame(
  outcome = theta_1_measures$Z_theta1_1,
  skill_proxy = theta_0_factor$residual_measures$Z_theta0_1,
  investment_proxy = I_0_factor$residual_measures$Z_I0_1,
  skill_instrument = theta_0_factor$residual_measures$Z_theta0_2,
  investment_instrument = I_0_factor$residual_measures$Z_I0_2
)

aw_data$proxy_interaction <-
  aw_data$skill_proxy * aw_data$investment_proxy

aw_data$instrument_interaction <-
  aw_data$skill_instrument *
  aw_data$investment_instrument
```

## Naive translog regression

```{r naive-aw}
naive_aw_fit <- lm(
  outcome ~
    skill_proxy +
    investment_proxy +
    proxy_interaction,
  data = aw_data
)

naive_aw_coef <- coef(naive_aw_fit)[
  c("skill_proxy", "investment_proxy", "proxy_interaction")
]
```

## Manual two-stage least squares

The first stage projects every endogenous regressor onto the excluded
alternative measures:

$$
\left[
\widetilde Z_{\theta,1},
\widetilde Z_{I,1},
\widetilde Z_{\theta,1}\widetilde Z_{I,1}
\right]
\quad\text{on}\quad
\left[
\widetilde Z_{\theta,2},
\widetilde Z_{I,2},
\widetilde Z_{\theta,2}\widetilde Z_{I,2}
\right].
$$

The second stage uses the fitted values from these first-stage regressions.

```{r aw-2sls}
instrument_set <- aw_data[
  c(
    "skill_instrument",
    "investment_instrument",
    "instrument_interaction"
  )
]

endogenous_set <- aw_data[
  c(
    "skill_proxy",
    "investment_proxy",
    "proxy_interaction"
  )
]

first_stage_fits <- lapply(
  endogenous_set,
  function(x) {
    lm(
      x ~
        skill_instrument +
        investment_instrument +
        instrument_interaction,
      data = cbind(
        x = x,
        instrument_set
      )
    )
  }
)

fitted_endogenous <- as.data.frame(
  lapply(first_stage_fits, fitted)
)

names(fitted_endogenous) <- c(
  "skill_hat",
  "investment_hat",
  "interaction_hat"
)

second_stage_data <- cbind(
  outcome = aw_data$outcome,
  fitted_endogenous
)

aw_iv_fit <- lm(
  outcome ~
    skill_hat +
    investment_hat +
    interaction_hat,
  data = second_stage_data
)

aw_iv_coef <- coef(aw_iv_fit)[
  c("skill_hat", "investment_hat", "interaction_hat")
]
```

## Compare the estimates

```{r compare-aw}
aw_comparison <- data.frame(
  Parameter = c(
    "Current skill",
    "Investment",
    "Skill-investment interaction"
  ),
  True = c(b1_true, b2_true, b3_true),
  Naive_OLS = as.numeric(naive_aw_coef),
  AW_style_IV = as.numeric(aw_iv_coef)
)

kable(aw_comparison, digits = 3)
```

The point of the comparison is not that this small simulation reproduces all
of AW. The point is to make the source of the correction visible:

1. the main measures contain the latent variables plus noise;
2. the noisy regressors are endogenous because their noise enters the error;
3. alternative measures remain correlated with the latent signal;
4. the alternative measures can therefore serve as internal instruments.

# 5. Where age-invariance enters

The repeated-measure IV above addresses measurement error. It does not by
itself connect the factor scale across ages.

AW additionally requires at least one suitable age-invariant measure for each
latent factor, or an equivalent set of overlapping measures. For an
age-invariant anchor:

$$
\mu_{\theta,t,m}=\mu_{\theta,0,m},
\qquad
\lambda_{\theta,t,m}=\lambda_{\theta,0,m}
\qquad
\text{for every }t.
$$

In the simulation, the first cognitive measure can be interpreted as such an
anchor because its loading is set to one at both periods:

```{r age-invariant-anchor}
age_invariant_example <- data.frame(
  Period = c("t = 0", "t = 1"),
  Anchor = c("Z_theta0_1", "Z_theta1_1"),
  Loading = c(1, 1),
  Interpretation = c(
    "Initial-period scale",
    "Same scale carried forward"
  )
)

kable(age_invariant_example, align = "lccc")
```

In an empirical application, this is a maintained measurement restriction.
It should not be confused with simply converting each test score into an
age-specific \(z\)-score.

# 6. What this notebook leaves out

This is intentionally smaller than the full AW implementation. It does not
include:

- multiple developmental stages;
- cognitive and noncognitive skills simultaneously;
- a complete investment policy function;
- all covariance restrictions used in the paper;
- the full sequential recovery of every measurement parameter;
- bootstrap inference across all estimation steps;
- the complete adult-outcome anchoring procedure.

The notebook is designed to make the logic of the estimator transparent:

$$
\text{CES benchmark}
\;\longrightarrow\;
\text{measurement error}
\;\longrightarrow\;
\text{internal IV}
\;\longrightarrow\;
\text{AW-style translog estimation}.
$$

```{r session-info, echo=FALSE}
sessionInfo()
```
