---
title: "A Hands-On Demonstration with the CHS Working Sample"
subtitle: "An executable R/Quarto introduction to skill-formation data"
author: "Child Skill Formation"
date: "2026-09-01"
engine: knitr
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    html-math-method: katex
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 4.5,
  fig.align = "center"
)

options(width = 100)
library(knitr)
```

# Purpose

This page is a **teaching demonstration**, not a replication of the full
Cunha--Heckman--Schennach (CHS) estimator. We use the Appendix A9-style
teaching extract created from the local NLSY79 files and make the first steps
of a skill-formation analysis executable:

1. define the working sample;
2. inspect the representative measures by developmental period;
3. audit missingness;
4. construct transparent, period-specific proxy measures; and
5. estimate a simple benchmark production equation.

The working sample is **2,219 children**: firstborn children in the CHS
white-sample definition who have all three child measure types somewhere in
the A9 periods and a complete maternal A9 measurement vector. The published
CHS reference sample is shown separately as $N=2{,}207$.

::: {.callout-warning}
## Important interpretation

The regressions below use observed proxies. They do **not** recover latent
factors, correct for measurement error, identify a causal investment effect,
or impose the full CHS/AW age-invariance and normalization framework.
:::

# 1. Load the teaching extract

The source files are generated by
[`make_chs_white_children.R`](../make_chs_white_children.R). The path search
makes the page usable when the working directory is either the repository
root or `estimation_demo/`.

```{r}
#| label: load-data

data_files <- c(
  period = "chs_white_firstborn_appendix_a9_period.csv",
  child = "chs_white_firstborn_appendix_a9_child.csv"
)

candidate_roots <- unique(c(
  getwd(),
  dirname(getwd()),
  "/content/child-skill-formation",
  "/content/Child Skill Formation"
))

find_data_file <- function(filename) {
  candidates <- unique(unlist(lapply(
    candidate_roots,
    function(root) c(
      file.path(root, "data", "derived", filename),
      file.path(root, filename)
    )
  )))

  existing <- candidates[file.exists(candidates)]
  if (length(existing) == 0L) {
    stop(
      "Could not find ", filename, ". In Colab, upload the two derived ",
      "CSV files or clone the repository."
    )
  }
  existing[1L]
}

period_path <- find_data_file(data_files[["period"]])
child_path <- find_data_file(data_files[["child"]])

period <- read.csv(period_path, stringsAsFactors = FALSE, check.names = FALSE)
child <- read.csv(child_path, stringsAsFactors = FALSE, check.names = FALSE)

cat("Period file:", period_path, "\n")
cat("Child file: ", child_path, "\n")
cat("Period rows:", format(nrow(period), big.mark = ","), "\n")
cat("Child rows: ", format(nrow(child), big.mark = ","), "\n")
```

# 2. Define the 2,219-child working sample

The restriction is implemented in code rather than typed into a table. Every
later result can therefore be traced back to the sample rule.

```{r}
#| label: define-working-sample

has_three_types <- as.logical(child$child_has_all_three_a9_types)
maternal_complete <- as.logical(child$maternal_a9_complete)

working_ids <- unique(child$child_id[has_three_types & maternal_complete])
working <- period[period$child_id %in% working_ids, ]
working <- working[order(working$child_id, working$period), ]

stopifnot(length(working_ids) == 2219L)

sample_counts <- data.frame(
  `Sample rule` = c(
    "Race code 3 and firstborn",
    "At least one age-eligible A9 child measure",
    "All three A9 child measure types",
    "Complete maternal A9 measurement vector",
    "Working sample: both child and maternal restrictions",
    "Published CHS reference sample"
  ),
  Children = c(
    nrow(child),
    sum(as.logical(child$child_has_any_a9_measure)),
    sum(has_three_types),
    sum(maternal_complete),
    length(working_ids),
    2207L
  ),
  check.names = FALSE
)

kable(sample_counts, align = c("l", "r"))
```

The two numbers that should be kept separate are:

- $2{,}219$: our transparent working sample under the joint observable
  restriction;
- $2{,}207$: the published CHS reference count.

The 12-child difference is not silently removed. Matching the published
likelihood sample exactly would require the authors' original prepared data
and any additional flags used in estimation.

# 3. What is measured in each developmental period?

The Appendix A9 construction chooses a representative measure for each factor
and period. The selected measure changes over the life cycle.

```{r}
#| label: a9-map

first_nonmissing_text <- function(x) {
  x <- x[!is.na(x) & nzchar(x)]
  if (length(x) == 0L) NA_character_ else x[1L]
}

period_labels <- c(
  `1` = "Birth",
  `2` = "1--2",
  `3` = "3--4",
  `4` = "5--6",
  `5` = "7--8",
  `6` = "9--10",
  `7` = "11--12",
  `8` = "13--14"
)

a9_map <- do.call(rbind, lapply(sort(unique(working$period)), function(p) {
  group <- working[working$period == p, ]
  data.frame(
    Period = p,
    `Age range` = unname(period_labels[as.character(p)]),
    `Cognitive measure` = first_nonmissing_text(group$cognitive_measure_name),
    `Noncognitive measure` = first_nonmissing_text(group$non_cognitive_measure_name),
    `Investment measure` = first_nonmissing_text(group$investment_measure_name),
    check.names = FALSE
  )
}))

kable(a9_map, align = "l")
```

For example, cognitive skill is represented by birth weight in the first
period, MSD in the second, PPVT in the third, and PIAT Mathematics from the
fourth period onward. This connects the latent factor $\theta_t$ to actual
variables without pretending that the observed measures are error-free.

# 4. Audit the signal available in the data

Before estimating anything, count the nonmissing observations by period. A
working sample can contain 2,219 children while a particular transition uses
fewer observations because a selected measure is missing in that period.

```{r}
#| label: missingness-table

measure_columns <- c(
  Cognitive = "cognitive_measure",
  Noncognitive = "non_cognitive_measure",
  Investment = "investment_measure"
)

coverage <- do.call(rbind, lapply(sort(unique(working$period)), function(p) {
  group <- working[working$period == p, ]
  result <- data.frame(
    Period = p,
    `Age range` = unname(period_labels[as.character(p)]),
    Children = nrow(group),
    check.names = FALSE
  )
  for (label in names(measure_columns)) {
    values <- group[[measure_columns[[label]]]]
    result[[paste(label, "N")]] <- sum(!is.na(values))
    result[[paste(label, "%")]] <- 100 * mean(!is.na(values))
  }
  result
}))

kable(
  coverage,
  digits = 1,
  align = "r",
  caption = "Availability of representative A9 measures in the working sample."
)
```

```{r}
#| label: missingness-plot
#| fig-cap: "Availability of the representative A9 measures in the 2,219-child working sample."

percent_columns <- c("Cognitive %", "Noncognitive %", "Investment %")
matplot(
  coverage$Period,
  coverage[, percent_columns],
  type = "b",
  pch = 16,
  lty = 1,
  lwd = 2,
  col = c("#234E70", "#A04E61", "#6B7F8C"),
  xlab = "Developmental period",
  ylab = "Nonmissing observations (%)",
  ylim = c(0, 105),
  xaxt = "n"
)
axis(1, at = coverage$Period, labels = coverage$Period)
legend(
  "bottomleft",
  legend = c("Cognitive", "Noncognitive", "Investment"),
  col = c("#234E70", "#A04E61", "#6B7F8C"),
  lty = 1,
  pch = 16,
  lwd = 2,
  bty = "n",
  horiz = TRUE
)
```

The plot is already teaching us something important: the sample-size
restriction is at the child level, while the usable estimation sample is
transition-specific.

# 5. Build transparent proxy measures

The selected variables have different units: ounces, test scores, frequency
responses, and behavior scores. For this first benchmark, we standardize each
selected measure **within period**:

$$
X^{z}_{it} = \frac{X_{it}-\bar X_t}{s_t}.
$$

This is a transparent teaching choice, not the CHS latent-factor
normalization and not AW's age-invariance solution.

```{r}
#| label: standardize-proxies

z_score <- function(x) {
  if (sum(!is.na(x)) < 2L) {
    return(rep(NA_real_, length(x)))
  }
  spread <- sd(x, na.rm = TRUE)
  if (is.na(spread) || spread == 0) {
    return(rep(NA_real_, length(x)))
  }
  (x - mean(x, na.rm = TRUE)) / spread
}

working$cognitive_z <- ave(
  working$cognitive_measure,
  working$period,
  FUN = z_score
)
working$noncognitive_z <- ave(
  working$non_cognitive_measure,
  working$period,
  FUN = z_score
)
working$investment_z <- ave(
  working$investment_measure,
  working$period,
  FUN = z_score
)

head(working[, c(
  "child_id", "period", "cognitive_measure", "cognitive_z",
  "non_cognitive_measure", "noncognitive_z",
  "investment_measure", "investment_z"
)], 8)
```

If we later replace this step with a factor model, the rest of the data
pipeline can remain largely the same.

# 6. Stack adjacent periods into transitions

The canonical production equation asks how current inputs predict next-period
skill:

$$
\theta_{i,t+1} = f_t(\theta_{i,t}, I_{i,t}, \ldots).
$$

Here we create one row per child and adjacent period, keeping the period label
so that the benchmark can allow the average production environment to differ
by developmental stage.

```{r}
#| label: build-transitions

make_transition <- function(data, current_period) {
  current <- data[data$period == current_period, c(
    "child_id", "cognitive_z", "noncognitive_z", "investment_z"
  )]
  names(current) <- c(
    "child_id", "cognitive_t", "noncognitive_t", "investment_t"
  )

  next_period <- data[data$period == current_period + 1L, c(
    "child_id", "cognitive_z"
  )]
  names(next_period) <- c("child_id", "cognitive_tp1")

  transition <- merge(current, next_period, by = "child_id")
  transition$period_t <- current_period
  transition
}

transitions <- do.call(rbind, lapply(1:7, function(p) {
  make_transition(working, p)
}))

transition_counts <- do.call(rbind, lapply(1:7, function(p) {
  group <- transitions[transitions$period_t == p, ]
  complete <- complete.cases(group[, c(
    "cognitive_t", "noncognitive_t", "investment_t", "cognitive_tp1"
  )])
  data.frame(
    Transition = paste(p, "-->", p + 1L),
    `Child pairs` = nrow(group),
    `Complete rows` = sum(complete),
    check.names = FALSE
  )
}))

kable(transition_counts, align = "lrr")
```

The next-period outcome is missing for some child-period observations, so the
regression sample is smaller than 2,219. The 2,219 figure defines the
child-level working sample; each regression uses the observations needed for
its own variables.

# 7. Estimate a simple benchmark production equation

The following regression is intentionally modest:

$$
\widehat{\mathrm{Cognitive}}_{i,t+1}
= \beta_{0,t}
+ \beta_1\mathrm{Cognitive}_{i,t}
+ \beta_2\mathrm{Noncognitive}_{i,t}
+ \beta_3\mathrm{Investment}_{i,t}
+ u_{i,t+1}.
$$

The period fixed effects absorb average differences across transitions. The
coefficients should be read as associations between proxy measures, not as
causal production-function parameters.

```{r}
#| label: benchmark-regression

regression_data <- transitions[complete.cases(transitions[, c(
  "cognitive_t", "noncognitive_t", "investment_t", "cognitive_tp1"
)]), ]

benchmark_fit <- lm(
  cognitive_tp1 ~ cognitive_t + noncognitive_t + investment_t + factor(period_t),
  data = regression_data
)

benchmark_coefficients <- summary(benchmark_fit)$coefficients[
  c("cognitive_t", "noncognitive_t", "investment_t"),
  , drop = FALSE
]
colnames(benchmark_coefficients) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")

cat("Estimation observations:", format(nrow(regression_data), big.mark = ","), "\n\n")
kable(
  benchmark_coefficients,
  digits = 3,
  caption = "Proxy skill-formation regression"
)
```

The coefficient on `investment_t` is the benchmark association between the
current investment proxy and next-period cognitive skill, conditional on the
current skill proxies and period effects. It is not yet the CHS investment
productivity parameter because the regressors remain noisy and potentially
endogenous.

# 8. Add the skill--investment interaction

To connect the code to the complementarity discussion, add an interaction:

$$
\widehat{\mathrm{Cognitive}}_{i,t+1}
= \cdots + \beta_3 I_{i,t}
+ \beta_4 I_{i,t}\times \mathrm{Cognitive}_{i,t}
+ u_{i,t+1}.
$$

In this proxy regression, $\beta_4$ is a direct/log-style interaction on the
chosen proxy scale. It is not automatically the dynamic complementarity of the
full multi-period technology.

```{r}
#| label: interaction-regression

interaction_fit <- lm(
  cognitive_tp1 ~ cognitive_t + noncognitive_t + investment_t +
    cognitive_t:investment_t + factor(period_t),
  data = regression_data
)

interaction_coefficients <- summary(interaction_fit)$coefficients[
  c("cognitive_t", "investment_t", "cognitive_t:investment_t"),
  , drop = FALSE
]
colnames(interaction_coefficients) <- c("Estimate", "Std. Error", "t value", "Pr(>|t|)")
kable(interaction_coefficients, digits = 3, caption = "Proxy interaction regression")
```

The implied investment association at three illustrative current-skill levels
is:

```{r}
#| label: marginal-investment-association

beta_investment <- coef(interaction_fit)[["investment_t"]]
beta_interaction <- coef(interaction_fit)[["cognitive_t:investment_t"]]

marginal_effects <- data.frame(
  `Current cognitive proxy` = c(-1, 0, 1),
  `Implied investment coefficient` = c(
    beta_investment - beta_interaction,
    beta_investment,
    beta_investment + beta_interaction
  ),
  check.names = FALSE
)

kable(marginal_effects, digits = 3, align = "rr")
```

This is a convenient classroom visualization of heterogeneity in investment
associations. To interpret it as a structural complementarity result, we would
still need a valid measurement model, an identified investment equation, and a
carefully specified dynamic technology.

# 9. What has—and has not—been estimated?

The notebook has now taken the data from a child-level sample rule to a
working regression:

$$
\text{raw NLSY files}
\rightarrow \text{A9 representative measures}
\rightarrow \text{2,219-child working sample}
\rightarrow \text{period transitions}
\rightarrow \text{proxy regression}.
$$

The next methodological steps would be:

- estimate latent cognitive and noncognitive factors from multiple measures;
- impose and examine the relevant scale and age-invariance restrictions;
- model measurement error rather than treating the proxy as the factor;
- address investment endogeneity with a clearly justified strategy; and
- estimate the multi-stage technology according to the chosen identification
  argument.

This page is useful as a first demonstration because every row and coefficient
is executable, while the distinction between a transparent proxy exercise and
the full CHS/AW estimator remains explicit.

## Running the page

From the repository root:

```bash
quarto render estimation_demo/chs_working_sample_demo_r.qmd
```

The same `.qmd` can be opened in RStudio or VSCode. In RStudio, use the
**Render** button; in VSCode, use the Quarto preview command.
