---
title: "How the CHS Likelihood Works"
subtitle: "Actual NLSY/CYA observations and a transparent likelihood decomposition"
author: "Child Skill Formation"
date: "2026-09-02"
engine: knitr
format:
  html:
    toc: true
    toc-depth: 1
    code-fold: false
    html-math-method: katex
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(knitr)
```

# 0. Load the actual NLSY/CYA data

This notebook no longer generates artificial test scores, HOME measures, or
adult outcomes. It loads the local extract of white first-born children and
uses their actual observations. Latent skills and investment are still
unobserved; that is precisely why they must be integrated out.

```{r}
#| label: load-actual-data

# Load the working-sample flags and the three observed-data files.
child <- read.csv("data/chs_white_firstborn_appendix_a9_child.csv")
period <- read.csv("data/chs_white_firstborn_appendix_a9_period.csv")
measure <- read.csv("data/chs_white_firstborn_measurements_long.csv")
adult <- read.csv("data/chs_adult_outcomes_2020.csv")
```

The Appendix A9 rule selects children observed in all three A9 blocks and
whose maternal measures are complete.

```{r}
#| label: select-working-sample

# Apply the previously constructed Appendix A9 sample restrictions.
keep <- child$child_has_all_three_a9_types & child$maternal_a9_complete
working_ids <- unique(child$child_id[keep])
measure <- measure[measure$child_id %in% working_ids, ]
cat("Appendix A9 working sample:", length(working_ids), "children\n")
```

We use one transition: ages 5--6 ($t$) to ages 7--8 ($t+1$). If a child has
more than one interview inside an age bin, the code averages the available
observations inside that bin. This is data preparation, not construction of
a latent-skill proxy.

```{r}
#| label: collapse-age-bins

measure_vars <- c("piat_math_raw", "piat_reading_recognition_raw", "ppvt_raw",
  "bpi_antisocial_raw", "bpi_headstrong_raw", "bpi_hyperactive_raw",
  "home_read_b", "home_read_c", "home_music_c", "home_total_z",
  "home_cognitive_z", "home_emotional_z")
use <- measure[measure$chs_period_approx %in% c(4, 5), ]
by_age <- aggregate(use[measure_vars], list(child_id = use$child_id,
  period = use$chs_period_approx), function(x) mean(x, na.rm = TRUE))
```

```{r}
#| label: make-transition-panel

# Put the two age bins on the same child-level row.
p4 <- by_age[by_age$period == 4, ]; p4$period <- NULL
p5 <- by_age[by_age$period == 5, ]; p5$period <- NULL
panel <- merge(p4, p5, by = "child_id", suffixes = c("_t", "_tp1"))
cat("Children observed in both age bins:", nrow(panel), "\n")
```

Maternal ASVAB and self-esteem measures form $Z_3$. Actual 2020 schooling,
wage income, and conviction outcomes form $Z_4$.

```{r}
#| label: attach-parent-adult

mvars <- c("maternal_asvab_arithmetic_reasoning", "maternal_asvab_word_knowledge",
  "maternal_asvab_paragraph_composition", "maternal_se_worth",
  "maternal_se_good_qualities", "maternal_se_positive_attitude")
mother <- period[period$period == 4, c("child_id", mvars)]
mother <- mother[!duplicated(mother$child_id), ]
panel <- merge(panel, mother, by = "child_id", all.x = TRUE)
panel <- merge(panel, adult, by = "child_id", all.x = TRUE)
```

For the main transition sample we require two cognitive measures, three
noncognitive measures, and three HOME indices at both dates. Missing PPVT and
adult outcomes are allowed in the general sample likelihood.

```{r}
#| label: analysis-sample

core_base <- c("piat_math_raw", "piat_reading_recognition_raw",
  "bpi_antisocial_raw", "bpi_headstrong_raw", "bpi_hyperactive_raw",
  "home_total_z", "home_cognitive_z", "home_emotional_z")
core <- c(paste0(core_base, "_t"), paste0(core_base, "_tp1"), mvars)
analysis <- panel[complete.cases(panel[core]), ]
cat("Complete two-period teaching panel:", nrow(analysis), "children\n")
```

Standardization only changes the units of each observed variable. It does
not average measures or replace the latent factors.

```{r}
#| label: standardize-observed-data

analysis$log_wage_2020 <- log1p(pmax(analysis$wage_income_2020, 0))
zvars <- c(paste0(measure_vars, "_t"), paste0(measure_vars[1:6], "_tp1"),
  mvars, "log_wage_2020")
for (v in zvars) {
  analysis[[paste0(v, "_z")]] <- as.numeric(scale(analysis[[v]]))
}
analysis$college4_2020 <- as.integer(analysis$highest_grade_2020 >= 8)
```

The individual-measure HOME variables are sparse because questionnaires
change with age. We retain the larger sample for the model and select one
fully observed child only when displaying every numerical contribution.

# 1. Put the entire CHS system on the page

Let $k\in\{C,N\}$ index cognitive and noncognitive skills. A compact version
of the full system is

$$
\begin{aligned}
Z_{1,k,t,j,i}
 &= \mu_{1,k,t,j}+\alpha_{1,k,t,j}\ln\theta_{k,t,i}
    +\varepsilon_{1,k,t,j,i},\\
Z_{2,t,j,i}
 &= \mu_{2,t,j}+\alpha_{2,t,j}\ln I_{t,i}
    +\varepsilon_{2,t,j,i},\\
Z_{3,k,j,i}
 &= \mu_{3,k,j}+\alpha_{3,k,j}\ln\theta_{k,P,i}
    +\varepsilon_{3,k,j,i},\\
Z_{4,j,i}
 &= g_j(\theta_{C,T,i},\theta_{N,T,i},X_i)+\varepsilon_{4,j,i},\\
\ln I_{t,i}
 &= q_t(\ln\boldsymbol\theta_{t,i},
         \ln\boldsymbol\theta_{P,i},Y_{t,i})+\eta_{I,t,i},\\
\theta_{k,t+1,i}
 &= \left[\sum_{\ell=1}^{5}\gamma_{k,t,\ell}
 X_{k,t,\ell,i}^{\varphi_{k,t}}\right]^{1/\varphi_{k,t}}
    \exp(\eta_{k,t+1,i}).
\end{aligned}
$$

The five CES inputs are current cognitive skill, current noncognitive skill,
current investment, maternal cognitive skill, and maternal noncognitive
skill. Measurement errors, investment shocks, production shocks, and the
initial latent-factor distribution complete the probability model.

## The same equations can lead to GMM or MLE

GMM selects implications of the equations, such as means, covariances, and
orthogonality restrictions:

$$
E[m_i(\psi)]=0,
\qquad
\widehat\psi_{GMM}
=\arg\min_\psi\bar m(\psi)'W\bar m(\psi).
$$

MLE specifies the joint densities and integrates over the unobserved state
$\mathcal X_i$:

$$
L_i(\psi)=\int p(Z_i,\mathcal X_i\mid X_i;\psi)\,d\mathcal X_i,
\qquad
\widehat\psi_{ML}=\arg\max_\psi\sum_i\log L_i(\psi).
$$

| | GMM | MLE |
|:--|:--|:--|
| Uses | Selected moments implied by the system | Entire specified joint density |
| Distribution | Does not require every density to be specified | Requires distributional assumptions or a flexible approximation |
| Latent factors | Recovered through covariance and higher-moment restrictions | Integrated out of the likelihood |
| Trade-off | More robust to density misspecification | Potentially more efficient, but computationally heavier |

This notebook follows the MLE factorization because it makes every
likelihood contribution visible. It is a pedagogical full-information
representation; it should not be confused with a claim that the original
CHS replication code is one black-box simultaneous optimizer.

# 2. Measurement system: identify signal from covariance

For three measures of one factor,

$$
Z_j=\lambda_j\theta+\epsilon_j,\qquad \lambda_1=1,
$$

conditional independence gives

$$
\begin{aligned}
\operatorname{Cov}(Z_1,Z_2)&=\lambda_2\operatorname{Var}(\theta),\\
\operatorname{Cov}(Z_1,Z_3)&=\lambda_3\operatorname{Var}(\theta),\\
\operatorname{Cov}(Z_2,Z_3)&=\lambda_2\lambda_3
\operatorname{Var}(\theta).
\end{aligned}
$$

Therefore the factor variance and loadings can be written directly in terms
of observed covariances:

$$
\operatorname{Var}(\theta)=\frac{S_{12}S_{13}}{S_{23}},\qquad
\lambda_2=\frac{S_{23}}{S_{13}},\qquad
\lambda_3=\frac{S_{23}}{S_{12}}.
$$

The next calculation uses actual PIAT Mathematics, PIAT Reading Recognition,
and PPVT scores at ages 5--6.

```{r}
#| label: cognitive-covariance

cognitive_t <- paste0(c("piat_math_raw_t", "piat_reading_recognition_raw_t",
  "ppvt_raw_t"), "_z")
S_C <- cov(analysis[cognitive_t], use = "complete.obs")
V_C <- S_C[1, 2] * S_C[1, 3] / S_C[2, 3]
lambda_C <- c(1, S_C[2, 3] / S_C[1, 3], S_C[2, 3] / S_C[1, 2])
sigma_C <- sqrt(pmax(diag(S_C) - lambda_C^2 * V_C, 0.05))
round(cbind(loading = lambda_C, error_sd = sigma_C), 3)
```

For example, the first covariance moment is

$$
m_{12}=S_{12}-\lambda_1\lambda_2\operatorname{Var}(\theta).
$$

```{r}
#| label: one-gmm-moment

# These are the three off-diagonal factor-model moments.
g12 <- S_C[1, 2] - lambda_C[1] * lambda_C[2] * V_C
g13 <- S_C[1, 3] - lambda_C[1] * lambda_C[3] * V_C
g23 <- S_C[2, 3] - lambda_C[2] * lambda_C[3] * V_C
round(c(g12 = g12, g13 = g13, g23 = g23), 8)
```

The same transparent covariance calculation is repeated for every observed
three-measure block. This loop only avoids repeating identical algebra; it
does not construct factor scores.

```{r}
#| label: measurement-blocks

blocks <- list(
  Ct = cognitive_t,
  Nt = paste0(c("bpi_antisocial_raw_t", "bpi_headstrong_raw_t", "bpi_hyperactive_raw_t"), "_z"),
  I = paste0(c("home_read_b_t", "home_read_c_t", "home_music_c_t"), "_z"),
  PC = paste0(mvars[1:3], "_z"), PN = paste0(mvars[4:6], "_z"),
  C1 = paste0(c("piat_math_raw_tp1", "piat_reading_recognition_raw_tp1", "ppvt_raw_tp1"), "_z"),
  N1 = paste0(c("bpi_antisocial_raw_tp1", "bpi_headstrong_raw_tp1", "bpi_hyperactive_raw_tp1"), "_z"))
```

```{r}
#| label: all-measurement-parameters

loadings <- errors <- list()
for (b in names(blocks)) {
  S <- cov(analysis[blocks[[b]]], use = "complete.obs")
  V <- S[1, 2] * S[1, 3] / S[2, 3]
  loadings[[b]] <- c(1, S[2, 3] / S[1, 3], S[2, 3] / S[1, 2])
  errors[[b]] <- sqrt(pmax(diag(S) - loadings[[b]]^2 * V, 0.05))
}
```

The $0.05$ floor prevents a finite-sample Heywood variance from making this
small teaching calculation undefined. A production application should
estimate admissible covariance parameters directly rather than clip them.

# 3. Investment and CES production densities

For the numerical decomposition, choose one actual child observed in all
displayed measures and adult outcomes.

```{r}
#| label: choose-actual-child

need <- c(unlist(blocks), "log_wage_2020_z", "college4_2020", "convicted_2020")
demo <- analysis[complete.cases(analysis[need]), ]
r <- demo[1, ]
cat("Displayed child ID:", r$child_id, "\n")
cat("Fully observed candidates:", nrow(demo), "\n")
```

The following values are **one integration node**, not observed factor
scores. The observed $Z$ values come from the selected child; the latent
state is varied and eventually integrated out.

```{r}
#| label: one-latent-node

x <- c(PC = 0.15, PN = -0.10, Ct = 0.20, Nt = -0.25,
  I = 0.10, C1 = 0.30, N1 = -0.15)
round(x, 2)
```

At this node, the investment equation supplies a density for latent
investment rather than for a HOME proxy:

$$
\ln I_i=\beta_0+\beta_C\ln\theta_{C,t,i}
+\beta_N\ln\theta_{N,t,i}+\beta_{PC}\ln\theta_{C,P,i}
+\beta_{PN}\ln\theta_{N,P,i}+\eta_{I,i}.
$$

```{r}
#| label: investment-density

beta_I <- c(intercept = 0.10, Ct = 0.20, Nt = 0.10, PC = 0.15, PN = 0.10)
mu_I <- beta_I[1] + sum(beta_I[-1] * x[c("Ct", "Nt", "PC", "PN")])
sd_I <- 0.55
ll_investment <- dnorm(x["I"], mean = mu_I, sd = sd_I, log = TRUE)
round(c(observed_node = x["I"], mean = mu_I, log_density = ll_investment), 4)
```

The cognitive and noncognitive production densities evaluate how close the
next latent skills are to the CES predictions.

Recall that the entries of `x` are in logs. Define

$$
c_t=x[\texttt{Ct}],\quad n_t=x[\texttt{Nt}],\quad
i_t=x[\texttt{I}],\quad p_C=x[\texttt{PC}],\quad
p_N=x[\texttt{PN}].
$$

The level inputs passed to the CES function are therefore

$$
X(x)=\left(e^{c_t},e^{n_t},e^{i_t},e^{p_C},e^{p_N}\right).
$$

The conditional mean of next-period **log cognitive skill** is

$$
\begin{aligned}
\mu_{C1}
&=\log\!\left[
\sum_{\ell=1}^{5}\gamma_{C,\ell}X_\ell(x)^{\phi_C}
\right]^{1/\phi_C}\\
&=\frac{1}{\phi_C}\log\!\left[
\gamma_{C,1}e^{\phi_Cc_t}
+\gamma_{C,2}e^{\phi_Cn_t}
+\gamma_{C,3}e^{\phi_Ci_t}
+\gamma_{C,4}e^{\phi_Cp_C}
+\gamma_{C,5}e^{\phi_Cp_N}
\right].
\end{aligned}
$$

With the candidate values used below, this becomes

$$
\mu_{C1}=\frac{1}{0.35}\log\!\left[
0.30e^{0.35c_t}+0.15e^{0.35n_t}+0.25e^{0.35i_t}
+0.20e^{0.35p_C}+0.10e^{0.35p_N}
\right].
$$

Likewise, the conditional mean of next-period **log noncognitive skill** is

$$
\mu_{N1}=\frac{1}{0.20}\log\!\left[
0.15e^{0.20c_t}+0.35e^{0.20n_t}+0.25e^{0.20i_t}
+0.10e^{0.20p_C}+0.15e^{0.20p_N}
\right].
$$

The code implements these two expressions literally: `exp()` converts the
log states to levels, the weighted power sum calculates the CES output, and
the outer `log()` returns to the log-skill scale used by `x["C1"]` and
`x["N1"]`.

```{r}
#| label: ces-at-node

gamma_C <- c(0.30, 0.15, 0.25, 0.20, 0.10); phi_C <- 0.35
gamma_N <- c(0.15, 0.35, 0.25, 0.10, 0.15); phi_N <- 0.20
inputs <- exp(x[c("Ct", "Nt", "I", "PC", "PN")])
mu_C1 <- log(sum(gamma_C * inputs^phi_C)^(1 / phi_C))
mu_N1 <- log(sum(gamma_N * inputs^phi_N)^(1 / phi_N))
ll_C1 <- dnorm(x["C1"], mu_C1, 0.45, log = TRUE)
ll_N1 <- dnorm(x["N1"], mu_N1, 0.55, log = TRUE)
```

```{r}
#| label: display-production-density

production <- data.frame(
  skill = c("Cognitive", "Noncognitive"),
  latent_next = x[c("C1", "N1")],
  CES_prediction = c(mu_C1, mu_N1),
  log_density = c(ll_C1, ll_N1))
kable(production, digits = 4)
```

# 4. One actual child's full conditional contribution

For fixed $\mathcal X_i$, the joint contribution factors as

$$
\begin{aligned}
p(Z_i,\mathcal X_i)
={}&p(\theta_{P,i},\theta_{t,i})
p(Z_{3,i}\mid\theta_{P,i})p(Z_{1,t,i}\mid\theta_{t,i})\\
&\times p(I_{t,i}\mid\theta_{t,i},\theta_{P,i})
p(Z_{2,t,i}\mid I_{t,i})\\
&\times p(\theta_{t+1,i}\mid\theta_{t,i},I_{t,i},\theta_{P,i})
p(Z_{1,t+1,i}\mid\theta_{t+1,i})\\
&\times p(Z_{4,i}\mid\theta_{t+1,i}).
\end{aligned}
$$

Each actual observed measure now contributes its own Gaussian log density.

```{r}
#| label: observed-measure-table

factor <- rep(names(blocks), each = 3)
meas <- data.frame(factor, measure = sub("_z$", "", unlist(blocks)))
meas$z <- as.numeric(unlist(r[unlist(blocks)]))
meas$lambda <- as.numeric(unlist(loadings[names(blocks)]))
meas$sigma <- as.numeric(unlist(errors[names(blocks)]))
meas$latent_node <- as.numeric(x[meas$factor])
meas$log_density <- dnorm(meas$z, meas$lambda * meas$latent_node,
  meas$sigma, log = TRUE)
```

```{r}
#| label: display-measure-contributions

kable(meas, digits = 3)
```

The adult block illustrates that $Z_4$ can mix continuous and discrete
outcomes. Here four-year college and conviction use probit links, while log
wage income uses a Gaussian density.

```{r}
#| label: adult-outcome-contribution

p_college <- pnorm(-0.30 + 0.80 * x["C1"] + 0.40 * x["N1"])
p_convict <- pnorm(-1.20 - 0.30 * x["C1"] - 0.70 * x["N1"])
ll_college <- dbinom(r$college4_2020, 1, p_college, log = TRUE)
ll_convict <- dbinom(r$convicted_2020, 1, p_convict, log = TRUE)
mu_wage <- 0.40 * x["C1"] + 0.20 * x["N1"]
ll_wage <- dnorm(r$log_wage_2020_z, mu_wage, 0.90, log = TRUE)
ll_adult <- ll_college + ll_convict + ll_wage
```

Finally add the initial-state, measurement, investment, production, and
adult-outcome pieces. Every displayed number is evaluated at the same latent
node for the same actual child.

```{r}
#| label: full-node-contribution

ll_initial <- sum(dnorm(x[c("PC", "PN", "Ct", "Nt")], 0, 1, log = TRUE))
measure_by_block <- aggregate(log_density ~ factor, data = meas, sum)
structural <- data.frame(factor = c("Initial state", "Investment function",
  "Production: C", "Production: N", "Adult outcomes"),
  log_density = c(ll_initial, ll_investment, ll_C1, ll_N1, ll_adult))
contributions <- rbind(measure_by_block, structural)
kable(contributions, digits = 4)
```

```{r}
#| label: conditional-total

conditional_log_density <- sum(contributions$log_density)
cat("Full conditional log density at this node:",
  round(conditional_log_density, 4), "\n")
```

# 5. CHS-style computation: integrate, then maximize

The node above is not an estimate. The observed-data likelihood integrates
over all possible latent states:

$$
L_i(\psi)=\int p(Z_i,\mathcal X_i;\psi)\,d\mathcal X_i.
$$

This integral is the central computational object. For every trial value of
$\psi$, the estimator evaluates a large complete-data likelihood at many
latent paths, integrates those paths for every child, sums the marginal log
likelihoods, and then changes $\psi$ and starts again.

$$
\boxed{
\psi
\longrightarrow
\left\{p(Z_i,\mathcal X_i^{(b)};\psi)\right\}_{i,b}
\longrightarrow
\left\{L_i(\psi)\right\}_i
\longrightarrow
\sum_i\log L_i(\psi)
}
$$

The code below uses sequential Monte Carlo. Drawing from the initial,
investment, and production densities makes the marginal likelihood the
average of the remaining measurement and adult-outcome densities.

```{r}
#| label: draw-initial-investment

set.seed(20260902); B <- 20000L
draw <- data.frame(PC = rnorm(B), PN = rnorm(B),
  Ct = rnorm(B), Nt = rnorm(B))
draw$I <- beta_I[1] + beta_I["Ct"] * draw$Ct + beta_I["Nt"] * draw$Nt +
  beta_I["PC"] * draw$PC + beta_I["PN"] * draw$PN + rnorm(B, 0, sd_I)
```

```{r}
#| label: draw-production

Xdraw <- exp(as.matrix(draw[c("Ct", "Nt", "I", "PC", "PN")]))
index_C <- rowSums(sweep(Xdraw^phi_C, 2, gamma_C, "*"))
index_N <- rowSums(sweep(Xdraw^phi_N, 2, gamma_N, "*"))
draw$C1 <- log(index_C^(1 / phi_C)) + rnorm(B, 0, 0.45)
draw$N1 <- log(index_N^(1 / phi_N)) + rnorm(B, 0, 0.55)
```

```{r}
#| label: integrate-measurements

ll_observed <- rep(0, B)
for (j in seq_len(nrow(meas))) {
  f <- meas$factor[j]
  ll_observed <- ll_observed + dnorm(meas$z[j],
    meas$lambda[j] * draw[[f]], meas$sigma[j], log = TRUE)
}
```

```{r}
#| label: integrate-adult-outcomes

pc <- pnorm(-0.30 + 0.80 * draw$C1 + 0.40 * draw$N1)
pv <- pnorm(-1.20 - 0.30 * draw$C1 - 0.70 * draw$N1)
ll_observed <- ll_observed + dbinom(r$college4_2020, 1, pc, log = TRUE)
ll_observed <- ll_observed + dbinom(r$convicted_2020, 1, pv, log = TRUE)
ll_observed <- ll_observed + dnorm(r$log_wage_2020_z,
  0.40 * draw$C1 + 0.20 * draw$N1, 0.90, log = TRUE)
```

```{r}
#| label: marginal-likelihood-number

# Log-mean-exp prevents numerical underflow.
a <- max(ll_observed)
log_L_i <- a + log(mean(exp(ll_observed - a)))
cat("Monte Carlo marginal log likelihood:", round(log_L_i, 4), "\n")
```

For the whole sample,

$$
\ell(\psi)=\sum_{i=1}^{n}\log L_i(\psi).
$$

An optimizer changes the measurement, investment, production, shock, and
initial-distribution parameters, recomputes all $L_i(\psi)$, and searches for
the parameter vector that maximizes this sum. The computational difficulty
is not any single density. It is repeatedly performing a high-dimensional
integration for every child while respecting normalizations and parameter
constraints.

# 6. The Agostinelli--Wiswall alternative

Agostinelli--Wiswall do not require an individual posterior assignment of
$\theta_i$ inside the production-function estimator. Their sequential
approach first links the measurement scales across ages and then substitutes
rescaled noisy measures into the structural equations. Alternative measures
of the same latent factor serve as instruments for the primary measure.

Suppose the rescaled primary measures are

$$
\widetilde Z_{\theta,t,1}=\ln\theta_t+e_{\theta,t,1},
\qquad
\widetilde Z_{I,t,1}=\ln I_t+e_{I,t,1}.
$$

Substituting them into a translog technology gives

$$
\begin{aligned}
\widetilde Z_{\theta,t+1,1}
={}&a_t+\gamma_{1,t}\widetilde Z_{\theta,t,1}
+\gamma_{2,t}\widetilde Z_{I,t,1}\\
&+\gamma_{3,t}\widetilde Z_{\theta,t,1}
                    \widetilde Z_{I,t,1}+u_{t+1}.
\end{aligned}
$$

The first skill and investment measures are endogenous regressors because
their measurement errors enter $u_{t+1}$. Under independent measurement
errors, second measures contain the same latent signals but are orthogonal to
those errors. This produces IV moments such as

$$
E\!\left[
Q_{i,t}
\left(
\widetilde Z_{\theta,t+1,1}
-X_{i,t}'\gamma_t
\right)
\right]=0,
$$

where a simple instrument vector is

$$
Q_{i,t}=\left(
1,
\widetilde Z_{\theta,t,2},
\widetilde Z_{I,t,2},
\widetilde Z_{\theta,t,2}\widetilde Z_{I,t,2}
\right)'.
$$

## A small IV calculation with the actual measures

The following code illustrates the computational skeleton. PIAT Mathematics
is the primary skill measure, PIAT Reading Recognition is its alternative
measure, and two different HOME indices play the analogous investment roles.
The same baseline transformation is applied to Mathematics at both ages so
that growth is not mechanically removed by period-specific standardization.

```{r}
#| label: aw-linked-scale

aw <- analysis
math_mu <- mean(aw$piat_math_raw_t); math_sd <- sd(aw$piat_math_raw_t)
aw$skill_t <- (aw$piat_math_raw_t - math_mu) / math_sd
aw$skill_tp1 <- (aw$piat_math_raw_tp1 - math_mu) / math_sd
aw$skill_iv <- as.numeric(scale(aw$piat_reading_recognition_raw_t))
aw$invest_t <- as.numeric(scale(aw$home_total_z_t))
aw$invest_iv <- as.numeric(scale(aw$home_cognitive_z_t))
```

```{r}
#| label: aw-iv-matrices

# Primary measures enter X; alternative measures enter Q.
y_aw <- aw$skill_tp1
X_aw <- cbind(1, aw$skill_t, aw$invest_t, aw$skill_t * aw$invest_t)
Q_aw <- cbind(1, aw$skill_iv, aw$invest_iv, aw$skill_iv * aw$invest_iv)
A_aw <- crossprod(Q_aw, X_aw)
b_aw <- crossprod(Q_aw, y_aw)
W_aw <- solve(crossprod(Q_aw))
gamma_aw <- solve(t(A_aw) %*% W_aw %*% A_aw,
  t(A_aw) %*% W_aw %*% b_aw)
```

```{r}
#| label: aw-iv-result

aw_result <- data.frame(
  parameter = c("Intercept", "Skill", "Investment", "Skill x Investment"),
  estimate = as.numeric(gamma_aw))
kable(aw_result, digits = 3)
```

This numerical result is not an AW replication. It illustrates the estimator's
logic with our extract. A proper application must estimate and test the
age-invariant measurement restrictions, use the paper's full instrument set,
allow for missing observations, and address economic endogeneity separately
from measurement error.

## What changes computationally?

| | CHS-style latent likelihood | Agostinelli--Wiswall |
|:--|:--|:--|
| Individual skill | Remains latent inside $L_i(\psi)$ | Replaced by a rescaled noisy measure |
| Measurement error | Handled by integrating the latent state | Handled with alternative measures as IVs |
| Production estimation | Transition density inside a large likelihood | Sequential IV/GMM production equation |
| Main computation | High-dimensional integration for each child | Matrix moments and recursive IV steps |
| Essential scale restriction | Factor normalization | Age-invariant measures linking periods |

The contrast is therefore not ``latent skills'' versus ``observed skills.''
Both approaches treat skills as latent. The distinction is whether the
estimator integrates latent histories or eliminates them from the estimating
equations using transformed measures and moment restrictions.

## What this notebook has and has not done

- It uses actual NLSY/CYA child, HOME, maternal, schooling, wage, and
  conviction observations.
- It displays every density block in a CHS-style likelihood and integrates
  the latent state for one actual child.
- It does not pretend that standardized observed measures are latent-factor
  scores.
- It does not yet reproduce the full CHS estimates. Full replication requires
  all periods, CHS's distributional specification, missing-data treatment,
  anchoring choices, endogeneity treatment, and constrained optimization.
- It now shows the computational skeleton of the Agostinelli--Wiswall
  sequential IV alternative, but does not claim to reproduce their estimates.
